```{r}
options(repos = c(CRAN = "https://cloud.r-project.org/"))

```

The aim of this project is to fit a model that predicts median house value from the other variables in the data set

Load the data and check the overall structure of the data
```{r}
data= read.csv("train.csv")
str(data)
``` 
This shows that all the variables were properly loaded into r, with data types integers and numerals. 

```{r}
summary(data)

```
```{r}
ggplot(data, aes(x = medianHouseValue)) + geom_histogram(binwidth = 50000, fill = "rosybrown", color = "black")

```

```{r}
ggplot(data, aes(x = housingMedianAge)) + geom_histogram(binwidth = 10, fill = "indianred4", color = "black")
```

```{r}
ggplot(data, aes(x = totalRooms)) + geom_histogram(binwidth = 1000, fill = "rosybrown4", color = "black")

```

```{r}
ggplot(data, aes(x = totalBedrooms)) + geom_histogram(binwidth = 200, fill = "maroon", color = "black")

```
```{r}
ggplot(data, aes(x = population)) + geom_histogram(binwidth = 500, fill = "brown4", color = "black")

```

```{r}
ggplot(data, aes(x = households)) + geom_histogram(binwidth = 500, fill = "darkolivegreen", color = "black")

```

```{r}
ggplot(data, aes(x = medianIncome)) + geom_histogram(binwidth = 1, fill = "darkseagreen4", color = "black")

```

```{r}
colors()
```


```{r}

install.packages("reshape2")
```

```{r}
library(reshape2)

data_scaled = as.data.frame(scale(data))


data_melted = melt(data_scaled)


ggplot(data_melted, aes(x = variable, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Box Plots for Scaled Variables", x = "Variables", y = "Scaled Values")


``` 
Variables, such as totalRooms and population, have noticeable outliers
The spread of the data varies across different variables. For example, medianIncome has a relatively narrow IQR compared to totalRooms


```{r}
library(maps)
library(mapdata)
map("state", region="california")
points(data$longitude, data$latitude
,col="blue", pch=16, cex=0.3)
```




```{r}
library(ggplot2)
# Example: Plotting house prices on a map with custom colors and smaller dots
ggplot(data, aes(x = longitude, y = latitude, color = medianHouseValue)) +
  geom_point(size = 1) +  # Adjust the size of the dots
  scale_color_gradient(low = "darkblue", high = "pink") +  # Change the color gradient
  labs(title = "House Prices by Location", x = "Longitude", y = "Latitude")

```



```{r}

install.packages("geosphere")
library(geosphere)
```

For the aspect of data explanation, we will scale the features to ensure that varying scales and metrics for the variables do not skew the interpretation of the model. 
```{r}
target="medianHouseValue"
features=setdiff(names(data), target)

```

```{r}

data=log1p(data+1)
centroid= colMeans(data[, c("longitude", "latitude")])
data$distance= distGeo(matrix(c(data$longitude,data$latitude), ncol=2), centroid)



data[features]= scale(data[features])
data$distance= scale(data$distance)


```
Created pairwise distances

```{r}
cor(data)
```


```{r}
model=lm(as.formula(paste(target, "~.")), data=data)
summary(model)
```

```{r}
test=read.csv("test.csv")
test_target="medianHouseValue"
test_features=setdiff(names(test), test_target)

centroid= colMeans(test[, c("longitude", "latitude")])

test$distance= distGeo(matrix(c(test$longitude,test$latitude), ncol=2), centroid)
test[[test_target]] <- log1p(test[[test_target]])
test[test_features]=scale(test[test_features])
test$distance= scale(test$distance)


predictions= predict(model, test)

```

```{r}
test[[test_target]]=expm1(test[[test_target]])
predictions= expm1(predictions)

```

```{r}
comparison=data.frame(Actual=test$medianHouseValue, predicted= predictions)
mae=mean(abs(comparison$Actual-comparison$predicted))
mae

```



```{r}
library(ggplot2)
ggplot(test, aes_string(x=comparison$Actual,y=comparison$predicted)) +
  geom_point() +
  geom_abline(intercept=0, slope=1, color = 'red') +
  labs(title="Actual vs Predicted", x="Actual Values", y="Predicted Values") + theme_minimal()
 
```

Tree-Based Models: Consider using models like Random Forest or Gradient Boosting, which can handle weak correlations better.